Type theory as a quotient inductive-recursive type

### 2024-12-31 tex/main.tex
Pen-and-paper presentation

\bibliographystyle{plainurl}% the mandatory bibstyle
\title{Type theory as a quotient inductive-recursive type}
\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{[funding]}
\author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
\authorrunning{L.-T.~Chen and T.-C.~Tsai}
\Copyright{Liang-Ting Chen and Tzu-Chun Tsai}
\ccsdesc[500]{Theory of computation~Type theory}
\keywords{inductive-inductive types, quotient inductive types, inductive-recursive types, substitution calculus, category with families}
\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

\supplement{The formal implementation is hosted at \url{https://github.com/genetsai95/DTT-QIIRT}.}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Fredrik Nordvall Forsberg, Hsiang-Shang Ko.}%optional

@@

\paragraph*{Contributions}
\begin{itemize}
  \item Exploring the potential use of quotient inductive-inductive-recursive types and definitions by overlapping patterns.
    In particular, we give
    \begin{itemize}
      \item a definition of parallel substitution calculus in type theory, whose type substitution rules are definitional and confluent, using quotient inductive-inductive-recursive types and definitions by overlapping patterns;
      \item a definition of type theory with Coquand universes and $\Pi$-types in type theory, extending the above substitution calculus.
    \end{itemize}
  \item Comparison with other definitions using quotient inductive-inductive types.
\end{itemize}

\paragraph*{Plan of the paper}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006}
\paragraph*{Single substitution calculus}
\cite{Kaposi2023,Kaposi2024a}
\paragraph*{Schemata of inductive types}
\cite{Altenkirch2016a,Kaposi2019}

\section{Metatheory and formalisation}
\cite{Pujet2022,Pujet2024}

\subsection{Inductive-recursive types}
\cite{Dybjer2003,Dybjer2000,Dybjer1999}
\subsection{Definitions by overlapping patterns}
\cite{Cockx2014}
\subsection{Rewrite rules}
\cite{Cockx2020,Cockx2021}
\paragraph*{Local confluence}
\paragraph*{Strong normalisation}

\section{Type theories as quotient inductive-inductive-recursive types} \label{sec:QIIRTs}
\subsection{Parallel substitution calculus} \label{subsec:SC}
\subsection{... with Coquand universes} \label{subsec:SC+U}
\cite{Coquand2013}
\subsection{... and \texorpdfstring{$\Pi$}{Π}-types} \label{subsec:SC+U+Pi}

>>>>

\bibliographystyle{plainurl}% the mandatory bibstyle
\title{Type theory as a quotient inductive-recursive type}
\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{Supported by the National Science and Technology Council of Taiwan under grant NSTC [funding]}
\author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
\authorrunning{L.-T.~Chen and T.-C.~Tsai}
\Copyright{Liang-Ting Chen and Tzu-Chun Tsai}
\ccsdesc[500]{Theory of computation~Type theory}
\keywords{inductive-inductive types, quotient inductive types, inductive-recursive types, substitution calculus, category with families}
\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

\supplement{The formal development is hosted at the GitHub repository: \url{https://github.com/genetsai95/DTT-QIIRT}.}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Fredrik Nordvall Forsberg, Hsiang-Shang Ko.}%optional

@@

\paragraph*{Contributions}
\begin{itemize}
  \item Exploring the potential use of quotient inductive-inductive-recursive types and definitions by overlapping patterns.
    In particular, we give in a type theory:
    \begin{itemize}
      \item a definition of parallel substitution calculus;
      \item a definition of type theory with Coquand universes and $\Pi$-types,
    \end{itemize}
   whose type substitution rules are definitional using quotient inductive-inductive-recursive types and definitions by overlapping patterns.
  \item Comparison with other definitions using quotient inductive-inductive types.
\end{itemize}

\subsection{Plan of the paper}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006}
\paragraph*{Single substitution calculus}
\cite{Kaposi2023,Kaposi2024a}
\paragraph*{Schemata of inductive types}
\cite{Altenkirch2016a,Kaposi2019}
\cite{Dybjer2003,Dybjer2000,Dybjer1999}

\section{Metatheory and formalisation}
\cite{Pujet2022,Pujet2024,Pujet2022a}

\subsection{Quotient inductive-inductive-recursive types}
\subsection{Definitions by overlapping patterns}
\cite{Cockx2014}
\subsection{Rewrite rules}
\cite{Cockx2020,Cockx2021}
\paragraph*{Local confluence}
\paragraph*{Strong normalisation}

\section{Type theories as quotient inductive-inductive-recursive types} \label{sec:QIIRTs}
\subsection{Parallel substitution calculus} \label{subsec:SC}
\[
\begin{alignedat}{10}
  \Ctx   & : \Set                   \\
  \Ty    & : \Ctx \to \Set          \\
  \Sub   & : \Ctx \to \Ctx \to \Set \\
  \Tm    & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Set \\
  \emptyctx & : \Ctx \\
  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\,\Gamma \to \Ctx \\
  [\blank]_{\Ty}\blank &: \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \\
  \emptysub & : \Sub\,\Gamma\,\emptyctx \\
  \blank,\blank & : \Sub\,\Gamma\,\Delta \to \Tm\,\Gamma\,([ \sigma ] A) \to \Sub\,\Gamma,(\Delta, A) \\
  \idS_{\blank} & : (\Gamma : \Ctx) \to \Sub\,\Gamma\,\Gamma \\
  \blank;\blank & : \Sub\,\Gamma\,\Delta \to \Sub\,\Delta\,\Theta \to \Sub\,\Gamma\,\Theta \\
  \pi_1 & : \Sub\,\Gamma\,(\Delta, A) \to \Sub\,\Gamma\,\Delta \\
  \pi_2 & : (\sigma : \Sub\,\Gamma\,(\Delta, A)) \to \Tm\,\Gamma\,([ \sigma ] A) \\
  [\blank]_{\Tm}\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma] A) \\
  [\blank] \blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma] A)
\end{alignedat}
\qquad
\begin{alignedat}{10}
          & [ \idS ]_{\Ty}\,A           && = A \\
          & [ \idS ]_{\Tm}\,t           && = t \\
          & [ \sigma ; \tau ]_{\Ty}\,A  && = [ \sigma ]_{\Ty}\,[ \tau ]_{\Ty} A \\
          & [ \sigma ; \tau ]_{\Tm}\,t  && = [ \sigma ]_{\Tm}\,[ \tau ]_{\Tm} t \\
          & [ \sigma ]_{\Tm}\,t         && = [ \sigma ]\,t & \text{otherwise}  \\
  [\idS]t & : [\,\idS\,]\,t             && = t \\
  [;]t    & : [\,\sigma ; \tau\,]\,t    && = [ \sigma ]\,[ \tau ]\,t \\
\end{alignedat}
\]

\subsection{... with Coquand universes} \label{subsec:SC+U}
\cite{Coquand2013}
\subsection{... and \texorpdfstring{$\Pi$}{Π}-types} \label{subsec:SC+U+Pi}
